EGGs are adhesive!

### 2025-03-12 calco/main.tex
term graph con equi

\begin{proposition}
Let $\Sigma$ be an algebraic signature. Then it holds
\begin{enumerate}
	\item if  $(i,j)\colon \mathcal{H}\to \mathcal{G}$ is a mono between  $(l, !_{V_\mathcal{G}})\colon \mathcal{G}\to \mathcal{G}^{\Sigma}$ and $(l', !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ in $\hyp_\Sigma$ and the latter is in $\tg$, then also the former is in $\tg$
	\item $\tg$ has equalizers, binary products and pullbacks and they are created by $I_\Sigma$.
\end{enumerate}
\end{proposition}

@@

\Cref{thm:slice-functors}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05}.
\begin{corollary}
	The category $\tg$ is quasiadhesive.
\end{corollary}

@@

As we have done in \Cref{sssect:hyp_alg_sign}, we can define the category of hypergraphs with equivalence labeled over an algebraic signature.

\begin{definition}
	Let $\Sigma = (O_\Sigma, \ari_\Sigma)$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to $\Sigma$.
	Then the \emph{hypergraph with equivalence} associated to $\Sigma$ is $R(\mathcal{G}^\Sigma)$ and the category of hypergraphs with 
	equivalence labeled over $\Sigma$ is the 
	slice category $\EqHyp_\Sigma = \EqHyp/L(\mathcal{G}^\Sigma)$.
\end{definition}

By \Cref{thm:slice_functors}, we can deduce the following.

\begin{proposition}
	$\EqHyp_\Sigma$ is $\reg (\EqHyp_\Sigma)$-adhesive.
\end{proposition}

%We can lift the adjunction given by $T$ and $L$ to $\EqHyp_\Sigma$ and $\hyp_\Sigma$.\todo{Non capisco se è uno svarione mio o ha senso}

We can define a functor $T_\Sigma: \EqHyps \to \hyps$ which sends each labeled hypergraph $l: \mathcal{G} \to R(\mathcal{G}^\Sigma)$ to $T_\Sigma(l) = T(l)$, and each morphism $h = (h_E, h_V, h_C)$ onto $T_\Sigma(h) = T(h)$. Such functor is faithful since so is $T$, hence we have the following property.

\begin{proposition}\label{prop:monos_in_eqhyps}
	A morphism $h$ between two objects of $\EqHyps$ is a mono if and only if $T(h)$ is a mono.
\end{proposition}

\begin{definition}
	Let $\Sigma$ be an algebraic signature.
	A labeled hypergraph with equivalence $l: \mathcal{G} \to R(\mathcal{G}^\Sigma)$ is a \emph{term graph with equivalence} if $t_{\mathcal{G}}$ is mono.
	We define category of term graphs with equivalence over $\Sigma$, denoted $\EqTG_{\Sigma}$, as the full subcategory of $\EqHyp_\Sigma$,
	and the corresponding inclusion functor $I_{\EqTG_\Sigma}$.
\end{definition}

\begin{proposition}
	If $l: \mathcal{G \to G}^\Sigma$ is a term graph, then $R(l)$ is a term graph with equivalence.
\end{proposition}

We have a functor $T_{\eq} : \EqTGs \to \tg$, defined to be $T_\Sigma \circ I_\eq$. Since both $T_\Sigma$ and $I_\eq$ are faithful, so is $T_\eq$.

If $h: \mathcal{H \to G}$ is a mono between $l': \mathcal{H} \to L(\mathcal{G}^\Sigma)$ and $l: \mathcal{G}\to L(\mathcal{G}^\Sigma)$, then, if $l\in\EqTGs$, $l'\in\EqTGs$ as well. This let us deduce what follows.

\begin{proposition}
	$\EqTGs$ has equalizers, and $I_{\EqTGs}$ creates them.
\end{proposition}

@@

@@

A straightforward application of \Cref{lim} now yields the following.
\begin{corollary}\label[corollary]{cor:pbapp}
	If $\X$ has pullbacks, then for every object $X$, the slice $\X/X$ has pullbacks too.
\end{corollary}

>>>>

\begin{proposition}\label{prop:tlim}
Let $\Sigma$ be an algebraic signature. Then it holds
\begin{enumerate}
	\item if  $(i,j)\colon \mathcal{H}\to \mathcal{G}$ is a mono between  $(l, !_{V_\mathcal{G}})\colon \mathcal{G}\to \mathcal{G}^{\Sigma}$ and $(l', !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ in $\hyp_\Sigma$ and the latter is in $\tg$, then also the former is in $\tg$
	\item $\tg$ has equalizers, binary products and pullbacks and they are created by $I_\Sigma$.
\end{enumerate}
\end{proposition}

@@

\Cref{thm:slice-functors}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05}.
\begin{corollary}\label{cor:term}
	The category $\tg$ is quasiadhesive.
\end{corollary}

@@

We are now going to generalize the work done in the previous section equipping term graphs with equivalence relation. First of all we need a notion of labelling for $\EqHyp$.

\begin{definition}
	Let $\Sigma$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to it. A \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, l)$ where $\mathcal{H}$ is an object of $\EqHyp$ and $l$ a morphism $T(\mathcal{H})\to \mathcal{G}^\Sigma$ of $\EqHyp$. A \emph{morphsim of labelled hypergraphs with equivalence} between $(\mathcal{H}, l)$ and $(\mathcal{H}', l')$ is an arrow $h\colon \mathcal{H}\to \mathcal{H}'$ such that $l= l'\circ T(\mathcal{H})$.

	We will denote the resulting category by $\EqHyp_\Sigma$.
\end{definition}

Let $(\mathcal{H}, l)$ be an object of $\EqHyp_\Sigma$, then, since $T$ has a right adjoint $R$ by \Cref{prop:forghyp}, $l\colon T(\mathcal{H})\to \mathcal{G}^\Sigma$ correponds to the arrow $(l, !_{Q_{\mathcal{H}}})\colon \mathcal{H}\to R(\mathcal{G}^\Sigma)$. It is immediate to see that this correspondence extends to an equivalence with the slice over $R(\mathcal{G}^\Sigma)$.

\begin{proposition}\label{prop:slice}
$\EqHyp_\Sigma$ is equivalent to $\EqHyp/R(\mathcal{G}^\Sigma)$.
\end{proposition}

$V_\Sigma\colon \EqHyp_\Sigma\to \EqHyp$ be the functor forgetting the labelling and let $\pbc_\Sigma$ be the class of morphisms in $\EqHyp_\Sigma$ whose image in $\EqHyp$ is in $\pbc$. From the second point of \Cref{thm:slice-functors,prop:forghyp,lim} we can deduce the following.

\begin{proposition}\label{prop:lim}The following facts hold true:
	\begin{enumerate}
		\item $\EqHyp_\Sigma$ has all colimits and all connected limits, which are created by the forgetful functor $V_{\Sigma}\colon \EqHyp_\Sigma\to \EqHyp$, in particular it has all equalizers and pullbacks;
		\item $\EqHyp_\Sigma$ is $\pbc_\Sigma$-adhesive.
	\end{enumerate}
\end{proposition}

Using \Cref{cor:mono1,cor:mono2} we immediately get the following result.

\begin{corollary}\label{prop:monos_in_eqhyps} Let $h=(h_{E}, h_V, h_Q)$ be an arrow in $\EqHyp_\Sigma$, then:
	\begin{enumerate}
		\item $h$ is mono if and only if $h_E$ and $h_V$ are injective;
		\item $h$ is a regular mono if and onky if $h_E$, $h_V$ and $h_Q$ are injective.
	\end{enumerate}
\end{corollary}

We can now easily define term graphs with equivalence.

\begin{definition}Let $\Sigma$ be an algebraic signature.
	An object $(\mathcal{H}, l)$ of $\EqHyp_\Sigma$ is a \emph{term graph with equivalence} if $l\colon T(\mathcal{H})\to \mathcal{G}^\Sigma$ is a term graph. We will denote by $\EqTG_{\Sigma}$ the full subcategory of $\EqHyp_\Sigma$ so defined and by $I_{\eq}$ the correspoding inclusion functor.
 \end{definition}

\noindent 
\parbox{9.5cm}{\begin{remark}\label{rem:obv}
Let $T_\Sigma$ be  $T\circ V_\Sigma$.   Notice that, by definition, $(\mathcal{H}, l)$ is in $\EqTG_{\Sigma}$ if and only if $T(\mathcal{H})$ is in $\tg$. Thus there exists a functor $S_\Sigma$ as in the diagram on the right.
\end{remark}}\hfill \parbox{4cm}{\xymatrix@R=15pt{\EqTG_{\Sigma} \ar[r]_{I_\eq}  \ar[d]_{S_\Sigma}& \EqHyp_\Sigma \ar[d]^{T_\Sigma} \\ \tg \ar[r]^{I_\Sigma}& \hyp_{\Sigma}}}

\begin{remark}\label{rem:t}
	Notice that, by \Cref{cor:limcolim} and \Cref{prop:lim}, the functor $T_\Sigma$ preserves all connected limits and all colimits.
\end{remark}

 The previous remark allows us to prove an analog of \Cref{prop:tlim}. We refer the reader to \Cref{proof:term} for details.

 \begin{restatable}{proposition}{trm}\label{prop:term}
$\EqTG_{\Sigma}$ has equalizers, binary products and pullbacks and they are created by $I_\eq$.
 \end{restatable}

\noindent
\parbox{11cm}{\hspace{15pt}Let now $\mathcal{T}$ be the class of morphism in $\EqTG_{\Sigma}$ whose image through $I_\eq$ is in $\pbc_\Sigma$ and whose image trhough $S_\Sigma$ is a regular mono in $\tg$.  Unwrapping the definition and using $\Cref{lem:reg}$, we have that a morphism $(h_E, h_V, h_Q)\colon (\mathcal{G},l)\to (\mathcal{H}, l')$ is in $\mathcal{T}$ if and only if all of its components are injections and the square on the right is a pullback.} \hfill \parbox{2cm}{\xymatrix{V_{\mathcal{G}}\ar@{>->}[r]^{h_V} \ar@{>>}[d]_{q_{\mathcal{H}}}& V_{\mathcal{H}} \ar@{>>}[d]^{q_{\mathcal{H}}}\\ Q_{\mathcal{G}} \ar@{>->}[r]^{h_Q}& Q_{\mathcal{H}}} }

\smallskip 
In particular, by \Cref{cor:term} and \Cref{lemma:stab}, $\mathcal{T}$ contains all isomorphisms, is closed under composition and decomposition and stable under pushout and pullbacks.

The following proposition is now an easy corollary of \Cref{prop:push,prop:lim,rem:obv}. We provide the details in \Cref{proof:tade}.

\begin{restatable}{proposition}{po}\label{prop:po}
$\EqTG_{\Sigma}$ has all $\mathcal{T}$.pushouts, which are created by $I_{\eq}$.
\end{restatable}

\begin{corollary}\label{cor:tade}
	$\EqTG_{\Sigma}$ is $\mathcal{T}$-adhesive.
\end{corollary}

@@

\trm*
\begin{proof}\label{proof:term}
	Let $F\colon \D\to \EqTG_{\Sigma}$ be the diagram of an equalizer, a binary product or of a pullback. By \Cref{prop:lim} we can consider a limiting cone $((\mathcal{L}, l), \{\pi_d\}_{d\in \D})$. By \Cref{rem:t} we know that $T_\Sigma$ preserves limits, thus by \Cref{rem:obv}  $(l, \{T_\Sigma(\pi_d)\}_{d\in \D})$ is limiting for $I_\Sigma\circ S_\Sigma$. By \Cref{prop:tlim} then $l$ is a term graph. We conclude, again by \Cref{rem:obv} that $(\mathcal{L}, l)$ is in $\EqTG_{\Sigma}$ and the thesis follows. 
\end{proof}

\po*
\begin{proof}\label{proof:tade}
	Suppose that the square on the left below is a pushout in $\EqHyp_\Sigma$, with $h$ in $\mathcal{T}$. Then, by \Cref{rem:t} the square on the right is a pushout and by the definition of $\mathcal{T}$ and \Cref{rem:obv}, $T_\Sigma(h)$ is a regular mono in $\tg$.
	\[\xymatrix{I_\eq(\mathcal{G}_0, l_0) \ar[r]^{k} \ar[d]_{h}& I_\eq(\mathcal{G}_1, l_1) \ar[d]^{p} &T_\Sigma (I_\eq(\mathcal{G}_0, l_0)) \ar[r]^{T_{\Sigma}(k)} \ar[d]_{T_{\Sigma}(h)}& T_\Sigma (I_\eq(\mathcal{G}_1, l_1)) \ar[d]^{T_\Sigma(p)}\\
I_\eq(\mathcal{G}_1, l_1) \ar[r]_{q}& I_\eq(\mathcal{H}, l) & T_\Sigma(I_\eq(\mathcal{G}_1, l_1)) \ar[r]_{T_\Sigma(q)}& T_\Sigma(I_\eq(\mathcal{H}, l))}\]

	We conclude using \Cref{prop:push,rem:obv}.
\end{proof}

@@

A straightforward application of \Cref{lim,colim} now yields the following.
\begin{corollary}\label[corollary]{cor:pbapp}
For eery object $\X$, $\X/X$ has all colimits and connected limits that $\X$ has. Moreover such limits and colimits are created by $\dom_X$.

In particular, if $\X$ has pullbacks, equalizers or pushouts, then for every object $X$, the slice $\X/X$ has such limits and colimits too.
\end{corollary}
